Hydras, Ordinals \& Co. \\ A library in Coq of entertaining formal mathematics

### 2021-10-04 doc/JFLA2022/paper.tex

\usepackage{varioref}
\newtheorem{todo}{To do}
\usepackage{amsfonts, afterpage}

\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}
\newcommand{\Hydras}{Hydras \& Co. }
\newcommand{\Coqcommunity}{\emph{Coq-community} }

@@

\newcommand{\TODO}[2][]{[\textcolor{red}{TODO (#1):} \emph{#2}]}

@@

\begin{itemize}
\item formalizations of mathematics in proof assistants, such as Coq, continually grow in size and scope
\begin{itemize}
\item the Mathematical Components library and the four-color theorem and the odd order theorem in Coq
\item the Mathlib library for Lean
\end{itemize}
\item to continue to be useful as examples and reusable libraries, formalizations must be well \emph{documented} and \emph{maintained} to ensure they work with new proof assistant releases. %\TODO[Pierre]{add evolution of styles of proof and formalization (not limited to Coq litterature)? } 
\item Lean and Mathematical Components have accompanying books, which are written mainly to onboard new library users
\item book authors faced with several design choices: literate documents, using tools to keep the book and the proof assistant synchronized
\end{itemize}

\subsection{Vision}
The \Hydras project, part of \Coqcommunity~\cite{CoqCommunity} aims to be an experimental platform for the collaborative development of commented libraries of formal proofs. \Coqcommunity is a community organization that we have founded in 2018 with two goals in mind: providing a solution for the long-term maintenance of interesting Coq packages, and working collaboratively on documentation projects. The \Hydras project demonstrates that these two goals are not independent: interesting Coq packages can become the basis for new documentation.
%
This umbrella project now includes evolved versions of the former Cantor and Additions libraries (under the new names of Hydra-battles and Addition-chains), the Ackermann sub-library, extracted from Russel O'Connor's Goedel library~\cite{OConnor05, Goedel} and a bridge to the Gaia library (by José Grimm~\cite{Gaia,grimm:hal-00911710}).
%
By following this approach of commenting interesting Coq packages, we provide new documentation content, that contributes to the diversity of the thriving Coq documentation ecosystem.

We call on the Coq users in the JFLA community and beyond to come and join us in this effort, by bringing new interesting projects which are worth presenting to Coq learners, \emph{a.k.a.} Coq users, and guiding them in their exploration.
%
We also always have project ideas to extend further our explorations and anyone is welcome to join the team by sending small or larger contributions through pull requests.
%
The current state of the project is already the result of such evolutions after several of us contributed project solutions and new proposals to the initial version of the first author.

Futhermore, contrary to traditionally published books, the ``book'' that forms part of this project is intended to be forever evolving. As new Coq formalization patterns and proof techniques appear, the book can be adapted to demonstrate their use (in case they fit well with our applications).
%
By using modern maintenance techniques such as continuous integration and deployment, we can ensure that this documentation stays up to date with the latest Coq releases. With Alectryon~\cite{alectryonpaper, alectryongithub}, we ensure that code and documentation are always in sync.

@@

The contrast between the simplicity of the statements above and the complexity of their proofs convinced us that it is a good theme for a commented library~\cite{HydraBattles} of formal proofs written for the \textit{Coq} proof assistant~\cite{Coq}.

@@

It was created in 2018, inspired by the \emph{Elm Community} organization~\cite{zimmermann:tel-02451322}.
%
Such ``Community Package Maintenance Organizations'' actually exist in many ecosystems as they avoid the very common problem of an important package becoming unmaintained as its author has moved on or disappeared~\cite{zimmermann2021grounded}.

In the case of Coq, this problem may be even more prevalent as many packages are created by PhD students, or by researchers for a specific paper, but their authors do not intend to maintain the package in the long-term.
%
The authors are however generally very open to having someone else who expresses interest in their work continue the maintenance.
%
\Coqcommunity makes that easy by providing a process for transferring or forking an unmaintained package, tooling for setting up good maintenance practices (such as continuous integration) and by making it possible for someone to take over a package without committing in the long-term (as maintainers who drop out can easily be replaced by some other volunteers).

Today, \Coqcommunity hosts over 50 projects maintained by over 30 maintainers.
%
The hosted projects come from a variety of origins.
%
Some had been maintained in the past by the Coq development team on behalf of the authors, but this meant that only minimal changes required to make the project build with new Coq versions were conducted.
%
Some were still maintained by their authors but they wanted help to keep them up-to-date.
%
Others were simply left unmaintained and were revived by their adoption in \Coqcommunity.

@@

\subsection{Documentation with Alectryon}

@@

Originally, we copied snippets from Coq sources into LaTeX manually, and recorded and inserted the corresponding outputs manually as well.  This approach is common, but brittle: changes to Coq definitions or lemmas had to be reflected in the book's sources, and we found multiple instances where the book and the Coq development had diverged. \TODO[Clément]{see FIXME cite alectryon for a discussion of various approaches to documentation of Coq developments}

@@

The transition happened gradually, over a few weeks: for each snippet of Coq code that was in the book, we had to take the following steps:

\begin{enumerate}
\item Mark the snippet in the Coq sources (we use special comments \texttt{(* begin snippet \textit{name} *)} … \texttt{(* end snippet \textit{name} *)})
\item Configure output display, using special Alectryon annotations to configure what should be shown (only the inputs, inputs and outputs, some steps of the proof but not all, some proof states at key moments in a proof, etc.)
\item Replace the copy-pasted inputs and output in LaTeX with an \texttt{input} command.
\end{enumerate}

Importing snippets into a LaTeX document was not one of the original use cases of Alectryon, so we had to extend it: the original Alectryon did not have support for exporting to LaTeX, and it was geared towards documenting individual source files (a style where code and prose are interleaved within the same file and code is documented in the same order as it is compiled).  We extended Alectryon to support our needs by implementing a LaTeX backend, and by programming it to generate individual snippet files, one per \texttt{snippet} comment block.  The later part was straightforward: all it took was to build a custom Alectryon \textit{driver}, a small (about 100 lines) Python program that leverages most of the Alectryon toolchain but defines a custom frontend that understands our snippet annotations (and otherwise exposes the exact same command line as Alectryon).

@@

\Hydras wants to be a bridge between scientific litterature (\emph{e.g.}~\cite{KP82, KS81, schutte} and proof assistant technology. For the mathematician, it may give a concrete view of the mathematical content, not only through full proofs, but also through significative computations: examples, functions associated with constructive proofs, etc. For the Coq user, it provides a consistent set of examples,
allowing to present and compare various formalization and proving techniques. It may be also used as a library, through opam packages.

It is also a medium sized library (more than 50K LOC), dependent on various tools and libraries of the Coq ecosystem. It may be also used to experiment new techniques of continuous maintenace of the code and its documentation.

@@

We plan to extend our libraries in two main directions.
The Gaia project~\cite{Gaia}, also maintained on \Coqcommunity contains a development by José Grimm~\cite{grimm:hal-00911710}. This library is dedicated  to the implementation in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics. It contains data structures which are compatible with our ordinal notations. 
Our plan is to build a  bridge between  the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems between both libraries.
Another direction would be to write in \textit{Coq} a formal proof of the original statement of Theorem~\ref{kp:thm2}, using O'Connor's formalization of Peano Arithmetic~\cite{Goedel}.

>>>>

\usepackage{varioref}
\newtheorem{todo}{To do}
\usepackage{amsfonts, afterpage}
\usepackage{xspace}

\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}
\newcommand{\Hydras}{Hydras \& Co.\xspace}
\newcommand{\Coqcommunity}{\emph{Coq-community\xspace} }

@@

\newcommand{\TODO}[2][]{[\textcolor{red}{TODO (#1):} \emph{#2}]}
\newcommand{\coq}{Coq\xspace}
\newcommand{\community}{Coq-community\xspace}
\newcommand{\gaia}{Gaia\xspace}
\newcommand{\alectr}{Alectryon\xspace}
\newcommand{\equations}{Equations\xspace}

@@

\begin{itemize}
\item formalizations of mathematics in proof assistants, such as \coq, continually grow in size and scope
\begin{itemize}
\item the Mathematical Components library and the four-color theorem and the odd order theorem in \coq
\item the Mathlib library for Lean
\end{itemize}
\item to continue to be useful as examples and reusable libraries, formalizations must be well \emph{documented} and \emph{maintained} to ensure they work with new proof assistant releases. %\TODO[Pierre]{add evolution of styles of proof and formalization (not limited to Coq litterature)? } 
\item Lean and Mathematical Components have accompanying books, which are written mainly to onboard new library users
\item book authors faced with several design choices: literate documents, using tools to keep the book and the proof assistant synchronized
\end{itemize}

\subsection{Vision}
The \Hydras project, part of \Coqcommunity\cite{CoqCommunity} aims to be an experimental platform for the collaborative development of commented libraries of formal proofs. \Coqcommunity is a community organization that we have founded in 2018 with two goals in mind: providing a solution for the long-term maintenance of interesting Coq packages, and working collaboratively on documentation projects. The \Hydras project demonstrates that these two goals are not independent: interesting Coq packages can become the basis for new documentation.

%
This umbrella project now includes evolved versions of the former Cantor and Additions libraries (under the new names of Hydra-battles and Addition-chains), the Ackermann sub-library, extracted from Russel O'Connor's Goedel library~\cite{OConnor05, Goedel} and a bridge to the Gaia library (by José Grimm~\cite{Gaia,grimm:hal-00911710}).
%
By following this approach of commenting interesting \coq packages, we provide new documentation content, that contributes to the diversity of the thriving \coq documentation ecosystem.

We call on the \coq users in the JFLA community and beyond to come and join us in this effort, by bringing new interesting projects which are worth presenting to \coq learners, \emph{a.k.a.} \coq users, and guiding them in their exploration.
%
We also always have project ideas to extend further our explorations and anyone is welcome to join the team by sending small or larger contributions through pull requests.
%
The current state of the project is already the result of such evolutions after several of us contributed project solutions and new proposals to the initial version of the first author.

Futhermore, contrary to traditionally published books, the ``book'' that forms part of this project is intended to be forever evolving. As new \coq formalization patterns and proof techniques appear, the book can be adapted to demonstrate their use (in case they fit well with our applications).
%
By using modern maintenance techniques such as continuous integration and deployment, we can ensure that this documentation stays up to date with the latest \coq releases. With {\alectr}~\cite{alectryonpaper, alectryongithub}, we ensure that code and documentation are always in sync.

@@

The contrast between the simplicity of the statements above and the complexity of their proofs convinced us that it is a good theme for a commented library~\cite{HydraBattles} of formal proofs written for the \coq proof assistant~\cite{Coq}.

@@

It was created in 2018, inspired by the {Elm Community} organization~\cite{zimmermann:tel-02451322}.
%
Such ``Community Package Maintenance Organizations'' actually exist in many ecosystems as they avoid the very common problem of an important package becoming unmaintained as its author has moved on or disappeared~\cite{zimmermann2021grounded}.

In the case of \coq, this problem may be even more prevalent as many packages are created by PhD students, or by researchers for a specific paper, but their authors do not intend to maintain the package in the long-term.
%
The authors are however generally very open to having someone else who expresses interest in their work continue the maintenance.
%

\Coqcommunity makes that easy by providing a process for transferring or forking an unmaintained package, tooling for setting up good maintenance practices (such as continuous integration) and by making it possible for someone to take over a package without committing in the long-term (as maintainers who drop out can easily be replaced by some other volunteers).

Today, \Coqcommunity hosts over 50 projects maintained by over 30 maintainers.
%
The hosted projects come from a variety of origins.
%
Some had been maintained in the past by the \coq development team on behalf of the authors, but this meant that only minimal changes required to make the project build with new \coq versions were conducted.
%
Some were still maintained by their authors but they wanted help to keep them up-to-date.
%
Others were simply left unmaintained and were revived by their adoption in \Coqcommunity.

@@

\subsection{Documentation with {\alectr}}

@@

Originally, we copied snippets from \coq sources into LaTeX manually, and recorded and inserted the corresponding outputs manually as well.  This approach is common, but brittle: changes to \coq definitions or lemmas had to be reflected in the book's sources, and we found multiple instances where the book and the \coq development had diverged. \TODO[Clément]{see FIXME cite alectryon for a discussion of various approaches to documentation of \coq developments}

@@

The transition happened gradually, over a few weeks: for each snippet of \coq code that was in the book, we had to take the following steps:

\begin{enumerate}
\item Mark the snippet in the \coq sources (we use special comments \texttt{(* begin snippet \textit{name} *)} … \texttt{(* end snippet \textit{name} *)})
\item Configure output display, using special {\alectr} annotations to configure what should be shown (only the inputs, inputs and outputs, some steps of the proof but not all, some proof states at key moments in a proof, etc.)
\item Replace the copy-pasted inputs and output in LaTeX with an \texttt{input} command.
\end{enumerate}

Importing snippets into a LaTeX document was not one of the original use cases of Alectryon, so we had to extend it: the original Alectryon did not have support for exporting to LaTeX, and it was geared towards documenting individual source files (a style where code and prose are interleaved within the same file and code is documented in the same order as it is compiled).  We extended Alectryon to support our needs by implementing a LaTeX backend, and by programming it to generate individual snippet files, one per \texttt{snippet} comment block.  The later part was straightforward: all it took was to build a custom Alectryon \textit{driver}, a small (about 100 lines) Python program that leverages most of the Alectryon toolchain but defines a custom frontend that understands our snippet annotations (and otherwise exposes the exact same command line as \alectr).

@@

\Hydras wants to be a bridge between scientific litterature (\emph{e.g.}~\cite{KP82, KS81, schutte}) and proof assistant technology. For the mathematician, it may give a concrete view of the mathematical content, not only through full proofs, but also through significative computations: examples, functions associated with constructive proofs, etc. For the Coq user, it provides a consistent set of examples,
allowing to present and compare various formalization and proving techniques. It may be also used as a library, through opam packages.

It is also a medium sized library (more than 50K LOC), dependent on various tools and libraries of the \coq ecosystem. It may be also used to experiment new techniques of continuous maintenance of the code and its documentation.

@@

We plan to extend our libraries in two main directions.

The Gaia project~\cite{Gaia}, also maintained on \Coqcommunity contains a development by José Grimm~\cite{grimm:hal-00911710}. This library is dedicated  to the implementation in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics. It contains data structures which are compatible with our ordinal notations.

Our plan is to build a  bridge between  the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems between both libraries.
Another direction would be to write in \coq a formal proof of the original statement of Theorem~\ref{kp:thm2}, using O'Connor's formalization of Peano Arithmetic~\cite{Goedel}.
